---
title: "Comparison of Classification and Clustering Algorithms on Glass Dataset Using R"
author: "Talha Hanif Butt"
output:
  pdf_document: default
  html_notebook: default
---

```{r}
library(caret)
library(mclust)
library(fpc)
library(cluster)
library(clusteval)
library(factoextra)
library(ggplot2)
library(kmed)
library(mlbench)
```

## Loading Glass Dataset
```{r}
# attach the Glass Identification dataset to the environment
data("Glass")
# rename the dataset
dataset <- Glass
```

## Partitioning Data for Validation
```{r}
# create a list of 80% of the rows inthe original dataset we can use for training
validation_index <- createDataPartition(dataset$Type, p=0.80, list=FALSE)
# select 20% of the data for validation
validation <- dataset[-validation_index,]
# use the remaining 80% of data to training and testing the models
dataset <- dataset[validation_index,]
```

## Getting Insights from Data
```{r}
# dimensions of dataset
dim(dataset)
```

```{r}
# list types for each attribute
sapply(dataset, class)
```

```{r}
# take a peek at the first 6 rows of the data
head(dataset)
```

```{r}
# list the levels for the class
levels(dataset$Type)
```

```{r}
# summarize the class distribution
percentage <- prop.table(table(dataset$Type)) * 100
cbind(freq=table(dataset$Type), percentage=percentage)
```

```{r}
# summarize attribute distributions
summary(dataset)
```

```{r}
# split input and output
x <- dataset[,1:9]
y <- dataset[,10]
```

```{r}
# boxplot for each attribute on one image
par(mfrow=c(1,5))
  for(i in 1:5) {
  boxplot(x[,i], main=names(Glass)[i])
}
```

```{r}
# boxplot for each attribute on one image
par(mfrow=c(1,4))
  for(i in 6:9) {
  boxplot(x[,i], main=names(Glass)[i])
}
```

```{r}
# barplot for class breakdown
plot(y)
```

```{r}
# scatterplot matrix
featurePlot(x=x, y=y, plot="ellipse")
```

```{r}
# box and whisker plots for each attribute
featurePlot(x=x, y=y, plot="box")
```

```{r}
ggplot(Glass, aes(RI,Na, color = Type)) + geom_point()
```

```{r}
ggplot(Glass, aes(Mg,Al, color = Type)) + geom_point()
```

```{r}
ggplot(Glass, aes(Si,K, color = Type)) + geom_point()
```

```{r}
# density plots for each attribute by class value
scales <- list(x=list(relation="free"), y=list(relation="free"))
featurePlot(x=x, y=y, plot="density", scales=scales)
```

## Applying Classification Algorithms
```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="cv", number=10)
metric <- "Accuracy"
```

```{r}
# kNN
set.seed(7)
fit.knn <- train(Type~., data=dataset, method="knn", metric=metric, trControl=control)
# SVM
set.seed(7)
fit.svm <- train(Type~., data=dataset, method="svmRadial", metric=metric, trControl=control)
# Random Forest
set.seed(7)
fit.rf <- train(Type~., data=dataset, method="rf", metric=metric, trControl=control)
```

## Comparison of the Classification Algorithms
```{r}
# summarize accuracy of models
results <- resamples(list(knn=fit.knn, svm=fit.svm, rf=fit.rf))
summary(results)
```

```{r}
# compare accuracy of models
dotplot(results)
```

## Insights from the best model
```{r}
# summarize Best Model
print(fit.rf)
```

```{r}
# estimate skill of Random Forest on the validation dataset
predictions <- predict(fit.rf, validation)
confusionMatrix(predictions, validation$Type)
```

## Applying Clustering Algorithms 
```{r}
# K-means
set.seed(20)
fit.kmeans <- kmeans(Glass[, 1:9], 7, nstart = 20)
# Hierarchical Agglomerative
set.seed(20)
d <- dist(Glass[,1:9], method = "euclidean") # distance matrix
fit.ha <- hclust(d, method="ward.D")
# K-Medoids Clustering
num <- as.matrix(Glass[,1:9])
mrwdist <- distNumeric(num, num, method = "mrw")
fit.kmedoids <- fastkmed(mrwdist, ncluster = 7, iterate = 50)
```

## Getting insights from Hierarchical Agglomerative Clustering
```{r}
# Cut tree into 4 groups
sub_grp <- cutree(fit.ha, k = 7)

# Number of members in each cluster
table(sub_grp)
## sub_grp
```

```{r}
plot(fit.ha, cex = 0.6)
rect.hclust(fit.ha, k = 7, border = 2:5)
```

```{r}
fviz_cluster(list(data = Glass[,1:9], cluster = sub_grp))
```

## Getting insights from K-Means Clustering
```{r}
table(fit.kmeans$cluster, Glass$Type)
```

```{r}
fit.kmeans$cluster <- as.factor(fit.kmeans$cluster)
ggplot(Glass, aes(RI, Na, color = fit.kmeans$cluster)) + geom_point()
```

```{r}
fit.kmeans$cluster <- as.factor(fit.kmeans$cluster)
ggplot(Glass, aes(Mg, Al, color = fit.kmeans$cluster)) + geom_point()
```

```{r}
fit.kmeans$cluster <- as.factor(fit.kmeans$cluster)
ggplot(Glass, aes(Si, K, color = fit.kmeans$cluster)) + geom_point()
```

## Getting insights from K-Medoids Clustering
```{r}
(fastiris <- table(fit.kmedoids$cluster, Glass[,10]))
```

```{r}
fit.kmedoids$cluster <- as.factor(fit.kmedoids$cluster)
ggplot(Glass, aes(RI, Na, color = fit.kmedoids$cluster)) + geom_point()
```

```{r}
fit.kmedoids$cluster <- as.factor(fit.kmedoids$cluster)
ggplot(Glass, aes(Mg, Al, color = fit.kmedoids$cluster)) + geom_point()
```

```{r}
fit.kmedoids$cluster <- as.factor(fit.kmedoids$cluster)
ggplot(Glass, aes(Si, K, color = fit.kmedoids$cluster)) + geom_point()
```

## Conclusion
With better accuracy and kappa measures, Random Forest has outperformed other competitors on Glass Dataset while Hierarchical Agglomerative Clustering is the winner when compared with K-Means and K-Medoids Clustering on Glass Dataset as it has clustered data better evident from the Cluster Plot and Cluster Dendrogram.